---
title: "EDA"
author: "Tory Norton"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F)
library(tidyverse)
library(sf)

library(patchwork)
library(scales)
library(scico)
library(RColorBrewer)
library(ggridges)

library(knitr)
theme_set(theme_bw())
heat_pal <- rev(brewer.pal(7, "RdBu"))
```


```{r}
segs <- read.table("DadeSegs")
colnames(segs) <- c("long1", "lat1", "long2", "lat2", "zip", "mystery")

# run in temporary environment to avoid creating a million objects
temp_env <- new.env()
source("zip_polys.R", local = temp_env)
zip_polys <- temp_env$zip_polys
zip_lines <- temp_env$lines
rm(temp_env)
```

```{r}
load("dadeRE.RData")
dade <- x
```


```{r}
bID <- filter(dade, pairID == "b")$propertyID

dade_a_matched0 <- dade %>% 
  filter(pairID == "a", propertyID %in% bID)
dade_a_matched <- dade_a_matched0 %>% 
  rename_with(
      ~ str_replace(
          .x,
          "\\d+$",
          ~ as.character(as.numeric(.x) + 1)
        ),
      matches("\\d+$")
    )

dade_wide <- dade %>% 
  anti_join(dade_a_matched, by = "outputID") %>% 
  select(-c(pairID, outputID)) %>% 
  left_join(select(dade_a_matched, -c(pairID, outputID)))

# nrow(dade_wide) + nrow(drop_na(dade_wide, amount3))
```





```{r}

dade_long <- dade_wide %>% 
  pivot_longer(ends_with(c("1", "2", "3"))) %>%
  mutate(sale = str_extract(name, "\\d"),
         name = str_extract(name, "[:alpha:]+")) %>% 
  pivot_wider() %>% 
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>% 
  drop_na(amount)

dade1 <- dade_long %>% 
  filter(sale == 1)

# total strata abundance
total_abundance <- dade1 %>%
  count(strata, name = "total_n") %>%
  mutate(total_abundance = total_n / sum(total_n))

# zipcode-level abundance
zip_abundance <- dade1 %>%
  mutate(zipcode = factor(zipcode, levels = zip_polys$zipcode, ordered = F),
         strata = factor(strata, levels = paste0("S", 1:5), ordered = T)) %>% 
  count(zipcode, strata, name = "zip_n") %>%
  complete(zipcode, strata, fill = list(zip_n = 0)) %>%   
  group_by(zipcode) %>% 
  mutate(zip_abundance = zip_n / sum(zip_n)) %>% 
  mutate(zipcode = as.numeric(as.character(zipcode)),
         strata = as.character(strata)) %>% 
  ungroup()

# compare
strata_sumry <- zip_abundance %>%
  left_join(total_abundance, by = "strata") %>%
  mutate(
    abundance_ratio = zip_abundance / total_abundance,
    abundance_diff = zip_abundance - total_abundance
  )
```

## Distribution of housing types


```{r}
strata_sumry %>% 
  group_by(strata) %>% 
  drop_na() %>% 
  summarize(proportion = unique(total_abundance),
            zips_without = sum(zip_abundance == 0)/n()
            # iqr = IQR(zip_abundance, na.rm = T)
            ) %>% 
  kable(digits = 3)
```

- Strata 3 is the most abundant housing type -- nearly 40% of homes in Dade county (judging by sales data)
  - does not occur in just 2.7% of residential zips
- Strata 5 is least abundant overall, and has highest proportion of zips where it does not occur at all

```{r}
p1 <- strata_sumry %>% 
  drop_na(zip_abundance) %>% 
  ggplot(aes(x = zip_abundance, y = strata, fill = strata, colour = strata)) +
  geom_density_ridges(alpha = 0.6, bw = 0.5) +
  scale_fill_viridis_d(aesthetics = c("color", "fill")) + 
  theme(legend.position = "none")

p2 <- strata_sumry %>% 
  drop_na(abundance_ratio) %>% 
  ggplot(aes(y = abundance_ratio, x = strata, fill = strata, color = strata)) +
  geom_violin(alpha = 0.5) +
  geom_hline(yintercept = 1, linetype = "dotted") +
  coord_flip() +
  scale_fill_viridis_d(aesthetics = c("color", "fill")) + 
  theme(legend.position = "none")

p1 + p2 + theme(axis.title.y = element_blank())
```

- S3 ratios are concentrated near 1 -- proportion of S3 housing is more constant across zips compared to other housing types
- S1 and S5 densities peak near zero -- do not occur in 11-12% of zips
- S4 and S5 ratio distributions are right-skewed -- tend to occur with high concentration in a small number of zips


```{r}
strata_sumry %>% 
  select(zipcode, strata, zip_abundance) %>% 
  pivot_wider(names_from = strata, values_from = zip_abundance) %>% 
  select(-zipcode) %>% 
  drop_na() %>% 
  cor() %>% 
  reshape2::melt() %>% 
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradientn(
    colours = rev(brewer.pal(7, "RdBu")),
    values = rescale(c(-0.5, 0, 1)),
    limits = c(-1, 1)
    )

```

- Housing types tend to cluster:
  - S1, S2 
  - S3, S4, S5


## Housing type by zip

```{r}
zip_sumry <- strata_sumry %>% 
  group_by(zipcode) %>% 
  summarise(
    zip_n = sum(zip_n), 
    flag = case_when(
      zip_n == 0 ~ "Non-residential",
      zip_n < 10 ~ "Small sample size"
      ))

zip_sumry %>% 
  drop_na(flag) %>% 
  left_join(zip_polys) %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(data = zip_polys) +
  geom_sf(aes(fill = flag)) +
  theme(legend.title = element_blank())
```


```{r}
downtown <- st_point(c(-80.1902, 25.7734)) %>% 
  st_sfc(crs = 4326) %>% 
  st_transform(st_crs(zip_polys$geometry))

p1 <- strata_sumry %>% 
  group_by(zipcode) %>% 
  slice_max(zip_abundance, with_ties = F, na_rm = T) %>% 
  ungroup() %>% 
  left_join(zip_polys) %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(data = zip_polys) + #grey background for NR zips
  geom_sf(aes(fill = strata)) +
  scale_fill_viridis_d() +
  labs(title = "Most abundant house type") +
  annotate("point", x = -80.1902, y = 25.7734, shape = 23, size = 2, fill = "red") +
  coord_sf(default_crs = sf::st_crs(4326)) +
  theme(axis.title = element_blank())

p2 <- strata_sumry %>% 
  group_by(zipcode) %>% 
  slice_max(abundance_ratio, with_ties = F, na_rm = T) %>% 
  ungroup() %>% 
  left_join(zip_polys) %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(data = zip_polys) + #bg
  geom_sf(aes(fill = strata)) +
  scale_fill_viridis_d() +
  labs(title = "Highest relative abundance") + 
  annotate("point", x = -80.1902, y = 25.7734, shape = 23, size = 2, fill = "red") +
  coord_sf(default_crs = sf::st_crs(4326)) +
  theme(axis.title = element_blank())


p1 + theme(legend.position = "none") + p2
```

- Most zips have highest proportion of S3 homes
- "Innner/outer city" pattern -- block of mostly S1 homes NW of downtown, surrounded by high prop of S5 homes 
- High abundance of S4 homes in suburban areas southwest of downtown


```{r}
strata_heatmap <- function(var, plot_strata = "S3", rng = NULL, midpoint = NULL, palette = rev(brewer.pal(7, "RdBu")), bg = "gray") {
  pal. <- palette
  if (!is.null(rng)){
    rng. <- rng
  } else {
    rng. <- range(pull(drop_na(filter(strata_sumry, strata %in% plot_strata), {{var}}), {{var}}))
  }
  if (!is.null(midpoint)) {
    mid. <- midpoint
  } else {
    mid. <- mean(rng.)
  }
  p <- expand.grid(strata = plot_strata, zipcode = zip_polys$zipcode) %>% 
  left_join(drop_na(filter(strata_sumry, strata %in% plot_strata), {{var}}), by = c("strata", "zipcode")) %>% 
  full_join(zip_polys, by = "zipcode") %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(aes(fill = {{var}})) +
  scale_fill_gradientn(
    colours = pal.,
    values = rescale(c(rng.[1], mid., rng.[2])),
    limits = rng.,
    na.value = bg
  )
  if (length(plot_strata >= 2)) {
    p <- p + facet_wrap(~ strata)
  }
  return(p)
}


strata_heatmap(zip_abundance, plot_strata = "S3", midpoint = 0.5, rng = c(0,1)) + 
  labs(title = "Proportion") +
  theme(legend.position = "left") +
strata_heatmap(abundance_ratio, "S3", midpoint = 1) +
  labs(title = "Relative abundance")

```

```{r}
strata_heatmap(var = abundance_ratio, plot_strata = c("S1", "S5"), midpoint = 1)
```

- some areas near coast have above average abundance of both S1 and S5 housing -- helps explain why S1 less negatively correlated with S5 than S3 and S4




---

## Sales activity by quarter

```{r}
year_activity <- dade_long %>% 
  filter(amount < 1e+07) %>% 
  group_by(year, strata) %>% 
  summarise(num_sales = n(), 
            avg_price = mean(amount))

year_activity %>% 
  ggplot(aes(x = year, y = num_sales, color = strata)) +
  geom_line()

year_activity %>% 
  ggplot(aes(x = year, y = avg_price, color = strata)) +
  geom_line() +
  scale_y_log10()
```

```{r}
total_price <- dade_long %>% 
  group_by(year) %>% 
  summarise(total_mean = mean(amount),
            total_med = median(amount))

zip_price <- dade_long %>% 
  group_by(year, zipcode) %>% 
  summarise(zip_mean = mean(amount),
            zip_med = median(amount))

price_sumry <- zip_price %>% 
  left_join(total_price) %>% 
  mutate(mean_ratio = zip_mean/total_mean,
         med_ratio = zip_med/total_med)

  
dade_long %>% 
  group_by(year, zipcode) %>% 
  summarise(avg_price = mean(amount)) %>% 
  mutate(year_bin = case_match(year, c(1991, 1992) ~ "1991-2",  c(1993, 1994) ~ "1993-4")) %>% 
  drop_na(year_bin) %>% 
  left_join(zip_polys) %>% 
  st_as_sf() %>% 
  ggplot(aes(fill = avg_price)) +
  geom_sf() + 
  facet_wrap(~year_bin)


plot_df <- price_sumry %>% 
  filter(year %in% c(1992, 1993, 1994))

rng <- range(plot_df$med_ratio)

plot_df %>% 
  left_join(zip_polys) %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(data = zip_polys) +
  geom_sf(aes(fill = med_ratio)) +
  facet_wrap(~year) +
  scale_fill_gradientn(
    colours = heat_pal,
    values = rescale(c(rng[1], 1, rng[2])),
    limits = rng,
    na.value = "gray80"
  )
```


---


```{r}
dade

```


