---
title: "EDA"
author: "Tory Norton"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F)
library(tidyverse)
library(sf)

library(patchwork)
library(scales)
library(scico)
library(RColorBrewer)
library(ggridges)

library(knitr)
theme_set(theme_light())
heatmap_pal <- rev(brewer.pal(7, "RdBu"))
```


```{r}
segs <- read.table("DadeSegs")
colnames(segs) <- c("long1", "lat1", "long2", "lat2", "zip", "mystery")

# run in temporary environment to avoid creating a million objects
temp_env <- new.env()
source("zip_polys.R", local = temp_env)
zip_polys <- temp_env$zip_polys
zip_lines <- temp_env$lines
rm(temp_env)
```


```{r}
lat_n <- 25.68
lat_s <- 25.3
  
andy_rect <- st_sf(
  geometry = st_sfc(
    st_polygon(list(matrix(
      c(-180, lat_n,
        180,  lat_n,
        180,  lat_s,
        -180, lat_s,
        -180, lat_n),
      ncol = 2,
      byrow = TRUE
    ))),
    crs = 4326
  )
) %>% 
  st_transform(st_crs(zip_polys))

andy_north <- st_sf(
  geometry = st_sfc(
    st_linestring(matrix(
      c(-180, lat_n,
        180,  lat_n),
      ncol = 2,
      byrow = TRUE
    )),
    crs = 4326
  )
) %>% 
  st_transform(st_crs(zip_polys))


zip_polys$hurricane <- as.logical(lengths(st_intersects(zip_polys, andy_rect)) > 0)

zip_polys %>% 
  ggplot() +
  geom_sf(aes(fill = hurricane))
```


```{r}
load("dadeRE.RData")
dade <- x
```


```{r}
bID <- filter(dade, pairID == "b")$propertyID

dade_a_matched0 <- dade %>% 
  filter(pairID == "a", propertyID %in% bID)
dade_a_matched <- dade_a_matched0 %>% 
  rename_with(
      ~ str_replace(
          .x,
          "\\d+$",
          ~ as.character(as.numeric(.x) + 1)
        ),
      matches("\\d+$")
    )

dade_wide <- dade %>% 
  anti_join(dade_a_matched, by = "outputID") %>% 
  select(-c(pairID, outputID)) %>% 
  left_join(select(dade_a_matched, -c(pairID, outputID)))

# nrow(dade_wide) + nrow(drop_na(dade_wide, amount3))

dade_long <- dade_wide %>% 
  pivot_longer(ends_with(c("1", "2", "3"))) %>%
  mutate(sale = str_extract(name, "\\d"),
         name = str_extract(name, "[:alpha:]+")) %>% 
  pivot_wider() %>% 
  mutate(date = ymd(paste(year, month, day, sep = "-"))) %>% 
  drop_na(amount)

dade1 <- dade_long %>% 
  filter(sale == 1)

# Add quarters
dade <- dade %>% 
  mutate(
    date1 = ymd(paste(year1, month1, day1, sep = "-")),
    date2 = ymd(paste(year2, month2, day2, sep = "-")),
    q1 = floor_date(date1, "quarter"),
    q2 = floor_date(date2, "quarter")
    ) %>% 
  filter(date1 <= date2)

dade_long <- dade_long %>% 
  mutate(q = floor_date(date, "quarter"))

```

```{r}
# total strata abundance
total_abundance <- dade1 %>%
  count(strata, name = "total_n") %>%
  mutate(total_abundance = total_n / sum(total_n))

# zipcode-level abundance
zip_abundance <- dade1 %>%
  mutate(zipcode = factor(zipcode, levels = zip_polys$zipcode, ordered = F),
         strata = factor(strata, levels = paste0("S", 1:5), ordered = T)) %>% 
  count(zipcode, strata, name = "zip_n") %>%
  complete(zipcode, strata, fill = list(zip_n = 0)) %>%   
  group_by(zipcode) %>% 
  mutate(zip_abundance = zip_n / sum(zip_n)) %>% 
  mutate(zipcode = as.numeric(as.character(zipcode)),
         strata = as.character(strata)) %>% 
  ungroup()

# compare
strata_sumry0 <- zip_abundance %>%
  left_join(total_abundance, by = "strata") %>%
  mutate(
    abundance_ratio = zip_abundance / total_abundance,
    abundance_diff = zip_abundance - total_abundance
  )

# remove zips with little data 
zip_sumry <- strata_sumry0 %>% 
  group_by(zipcode) %>% 
  summarise(
    zip_n = sum(zip_n), 
    flag = case_when(
      zip_n == 0 ~ "No data",
      zip_n < 10 ~ "Small sample size"
      )) %>% 
  left_join(select(as.data.frame(zip_polys), -geometry))

flagged_zips <- zip_sumry$zipcode[!is.na(zip_sumry$flag)]
strata_sumry <- strata_sumry0 %>% 
  filter(!zipcode %in% flagged_zips)
```

```{r}
year_activity <- dade_long %>% 
  filter(amount < 1e+07) %>% 
  group_by(year, strata) %>% 
  summarise(num_sales = n(), 
            avg_price = mean(amount))
```



### Test color palettes

```{r}
#map
m <- strata_sumry %>% 
  group_by(zipcode) %>% 
  slice_max(abundance_ratio, with_ties = F, na_rm = T) %>% 
  ungroup() %>% 
  left_join(zip_polys) %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(data = zip_polys) + #grey background for NR zips
  geom_sf(aes(fill = strata))
#lineplot
l <- year_activity %>% 
  ggplot(aes(x = year, y = num_sales, color = strata)) +
  geom_line()

library(gridExtra)

# pal1 <- c("#08306B", "#2171B5", "#4292C6", "#6BAED6", "#9ECAE1")
# pal2 <- c("#0B3C49", "#146C73", "#2A9D8F", "#6FC2B0", "#BFE5D6")
pal3 <- c("#1B3A4B", "#285C73", "#3E8E9E", "#73BFB8", "#A9D6CB")
pal4 <- c("#4A1486", "#6A51A3", "#807DBA", "#8DA0CB", "#A6BDD7")
pal5 <- viridisLite::viridis(5, begin = 0.05, end = 0.9)
pal_metal <- c(
  "#5A3A1B",  # bronze
  "#8C6D46",  # lighter bronze
  "#B8B8B8",  # silver
  "#D4AF37",  # gold
  "#F1D97A"   # light gold
)
pal_forest <- c( #UGLY!
  "#5E3C2C",
  "#7F5539",
  "#588157",
  "#3A7D44",
  "#1B4332"
)
pal6 <- c(
  "#2B2D6E",
  "#3F5AA6",
  "#4C8ED9",
  "#5FB5B2",
  "#8ED1C6"
)
pal7 <- c(
  "#2F3E46",
  "#354F52",
  "#52796F",
  "#84A98C",
  "#A4C3B2"
)
pal8 <- c(
  "#3B0F70",
  "#8C2981",
  "#DE4968",
  "#F66E5B",
  "#FE9F6D"
)

library(colorspace)


pal_test <- sequential_hcl(
  5,
  palette = "Viridis"
)

m + scale_fill_manual(values = pal_test)
l + scale_color_manual(values = pal_test)

pal_strata <- sequential_hcl(
  5,
  palette = "Viridis"
)
```


Problems:
- 45 rows where date1 > date2
  - just switch dates? switch amounts? drop?
- 2 sales with pairID b have no match in a


## Distribution of housing types


```{r}
strata_sumry %>% 
  group_by(strata) %>% 
  drop_na() %>% 
  summarize(proportion = unique(total_abundance),
            zips_without = sum(zip_abundance == 0)/n()
            # iqr = IQR(zip_abundance, na.rm = T)
            ) %>% 
  kable(digits = 3)
```

- Strata 3 is the most abundant housing type -- nearly 40% of homes in Dade county (judging by sales data)
  - does not occur in just 2.7% of residential zips
- Strata 5 is least abundant overall, and has highest proportion of zips where it does not occur at all

```{r}
p1 <- strata_sumry %>% 
  drop_na(zip_abundance) %>% 
  ggplot(aes(x = zip_abundance, y = strata, fill = strata, colour = strata)) +
  geom_density_ridges(alpha = 0.6, bw = 0.5) +
  scale_fill_viridis_d(aesthetics = c("color", "fill")) + 
  theme(legend.position = "none")

p2 <- strata_sumry %>% 
  drop_na(abundance_ratio) %>% 
  ggplot(aes(y = abundance_ratio, x = strata, fill = strata, color = strata)) +
  geom_violin(alpha = 0.5) +
  geom_hline(yintercept = 1, linetype = "dotted") +
  coord_flip() +
  scale_fill_viridis_d(aesthetics = c("color", "fill")) + 
  theme(legend.position = "none")

p1 + p2 + theme(axis.title.y = element_blank())
```

- S3 ratios are concentrated near 1 -- proportion of S3 housing is more constant across zips compared to other housing types
- S1 and S5 densities peak near zero -- do not occur in 11-12% of zips
- S4 and S5 ratio distributions are right-skewed -- tend to occur with high concentration in a small number of zips


```{r}
strata_sumry %>% 
  select(zipcode, strata, zip_abundance) %>% 
  pivot_wider(names_from = strata, values_from = zip_abundance) %>% 
  select(-zipcode) %>% 
  drop_na() %>% 
  cor() %>% 
  reshape2::melt() %>% 
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradientn(
    colours = rev(brewer.pal(7, "RdBu")),
    limits = c(-1, 1)
    )

```

- Housing types tend to cluster:
  - S1, S2 
  - S3, S4, S5


## Housing type by zip

```{r}
# zip_sumry <- strata_sumry %>% 
#   group_by(zipcode) %>% 
#   summarise(
#     zip_n = sum(zip_n), 
#     flag = case_when(
#       zip_n == 0 ~ "Non-residential",
#       zip_n < 10 ~ "Small sample size"
#       ))

zip_sumry %>% 
  drop_na(flag) %>% 
  left_join(zip_polys) %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(data = zip_polys) +
  geom_sf(aes(fill = flag)) +
  theme(legend.title = element_blank())
```


```{r}
downtown <- st_point(c(-80.1902, 25.7734)) %>% 
  st_sfc(crs = 4326) %>% 
  st_transform(st_crs(zip_polys$geometry))

p1 <- strata_sumry %>% 
  group_by(zipcode) %>% 
  slice_max(zip_abundance, with_ties = F, na_rm = T) %>% 
  ungroup() %>% 
  left_join(zip_polys) %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(data = zip_polys) + #grey background for NR zips
  geom_sf(aes(fill = strata)) +
  scale_fill_viridis_d() +
  labs(title = "Most abundant house type") +
  annotate("point", x = -80.1902, y = 25.7734, shape = 23, size = 2, fill = "red") +
  coord_sf(default_crs = sf::st_crs(4326)) +
  theme(axis.title = element_blank())

p2 <- strata_sumry %>% 
  group_by(zipcode) %>% 
  slice_max(abundance_ratio, with_ties = F, na_rm = T) %>% 
  ungroup() %>% 
  left_join(zip_polys) %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(data = zip_polys) + #bg
  geom_sf(aes(fill = strata)) +
  scale_fill_viridis_d() +
  labs(title = "Highest relative abundance") + 
  annotate("point", x = -80.1902, y = 25.7734, shape = 23, size = 2, fill = "red") +
  coord_sf(default_crs = sf::st_crs(4326)) +
  theme(axis.title = element_blank())


p1 + theme(legend.position = "none") + p2
```

- Most zips have highest proportion of S3 homes
- "Innner/outer city" pattern -- block of mostly S1 homes NW of downtown, surrounded by high prop of S5 homes 
- High abundance of S4 homes in suburban areas southwest of downtown


```{r}
p_tot <- filter(total_abundance, strata == "S3") %>% pull(total_abundance)


strata_sumry %>% 
  filter(strata == "S3") %>% 
  full_join(zip_polys, by = "zipcode") %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(aes(fill = zip_n)) +
  scale_fill_distiller(
    palette = "RdBu", 
    na.value = "grey"
    # labels = scales::percent,
    # limits = c(0, 0.8)
    ) +
  labs(
    title = "Number of S3 Homes",
    fill = ""
  )

rng <- filter(strata_sumry, strata == "S3")$zip_abundance %>% range(na.rm = T)

strata_sumry %>% 
  filter(strata == "S3") %>% 
  full_join(zip_polys, by = "zipcode") %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(aes(fill = zip_abundance)) +
  scale_fill_distiller(
    palette = "RdBu", 
    na.value = "grey", 
    labels = scales::percent,
    values = rescale(c(rng[1], 0.5, rng[2]))
    # limits = c(0, 0.8)
    ) +
  labs(
    title = "Proportion of S3 Homes",
    fill = ""
  )


```



---

## Sales activity by quarter

```{r}
year_activity %>% 
  ggplot(aes(x = year, y = num_sales, color = strata)) +
  geom_line()

year_activity %>% 
  ggplot(aes(x = year, y = avg_price, color = strata)) +
  geom_line() +
  scale_y_log10()
```

```{r}
total_price <- dade_long %>% 
  group_by(year) %>% 
  summarise(total_mean = mean(amount),
            total_med = median(amount))

zip_price <- dade_long %>% 
  group_by(year, zipcode) %>% 
  summarise(zip_mean = mean(amount),
            zip_med = median(amount))

price_sumry <- zip_price %>% 
  left_join(total_price) %>% 
  mutate(mean_ratio = zip_mean/total_mean,
         med_ratio = zip_med/total_med)

  
dade_long %>% 
  group_by(year, zipcode) %>% 
  summarise(avg_price = mean(amount)) %>% 
  mutate(year_bin = case_match(year, c(1991, 1992) ~ "1991-2",  c(1993, 1994) ~ "1993-4")) %>% 
  drop_na(year_bin) %>% 
  left_join(zip_polys) %>% 
  st_as_sf() %>% 
  ggplot(aes(fill = avg_price)) +
  geom_sf() + 
  facet_wrap(~year_bin)


plot_df <- price_sumry %>% 
  filter(year %in% c(1992, 1993, 1994))

rng <- range(plot_df$med_ratio)

plot_df %>% 
  left_join(zip_polys) %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf(data = zip_polys) +
  geom_sf(aes(fill = med_ratio)) +
  facet_wrap(~year) +
  scale_fill_gradientn(
    colours = heat_pal,
    values = rescale(c(rng[1], 1, rng[2])),
    limits = rng,
    na.value = "gray80"
  )
```


---


```{r}
andrew_qtr <- floor_date(ymd("1992-08-24"), "quarter")
  
qtr_activity <- dade_long %>%
  filter(amount < 1e+07) %>%
  group_by(q, strata) %>%
  summarise(
    num_sales = n(), 
    avg_price = mean(amount)
    ) %>% 
  mutate(S3 = strata == "S3")

qtr_activity %>% 
  ggplot(aes(x = q, y = num_sales, color = strata, alpha = S3)) +
  geom_line(linewidth = 0.7) +
  scale_alpha_manual(values = c(0.5, 1)) +
  scale_color_manual(values = pal_strata) +
  guides(alpha = "none") +
  geom_vline(xintercept = andrew_qtr, color = "red", linetype = "dashed")

qtr_activity %>% 
  ggplot(aes(x = q, y = avg_price, color = strata, alpha = S3)) +
  geom_line(linewidth = 0.7) +
  scale_y_log10() +
  scale_color_manual(values = pal_strata) +
  scale_alpha_manual(values = c(0.5, 1)) +
  guides(alpha = "none") +
  geom_vline(xintercept = andrew_qtr, color = "red", linetype = "dashed")

```

```{r}
qtr_activity_storm <- dade_long %>%
  left_join(select(zip_sumry, zipcode, hurricane)) %>% 
  filter(amount < 1e+07) %>%
  group_by(q, strata, hurricane) %>%
  summarise(
    num_sales = n(), 
    avg_price = mean(amount),
    med_price = median(amount)
    ) %>% 
  mutate(S3 = strata == "S3") %>% 
  ungroup()


qtr_activity_storm %>% 
  filter(S3) %>% 
  ggplot(aes(x = q, y = num_sales, color = hurricane)) +
  geom_line(linewidth = 0.7) +
  # scale_color_manual(values = pal_strata) +
  guides(alpha = "none") +
  geom_vline(xintercept = andrew_qtr, color = "red", linetype = "dashed") +
  scale_color_manual(values = c("steelblue2", "orange3"))
qtr_activity_storm %>% 
  filter(S3) %>% 
  ggplot(aes(x = q, y = avg_price, color = hurricane)) +
  geom_line(linewidth = 0.7) +
  # scale_color_manual(values = pal_strata) +
  guides(alpha = "none") +
  geom_vline(xintercept = andrew_qtr, color = "red", linetype = "dashed") +
  scale_color_manual(values = c("steelblue2", "orange3"))
qtr_activity_storm %>% 
  filter(S3) %>% 
  ggplot(aes(x = q, y = med_price, color = hurricane)) +
  geom_line(linewidth = 0.7) +
  # scale_color_manual(values = pal_strata) +
  guides(alpha = "none") +
  geom_vline(xintercept = andrew_qtr, color = "red", linetype = "dashed") +
  scale_color_manual(values = c("steelblue2", "orange3"))

```

---

## Modeling


```{r}
dade <- dade %>%
  mutate(log_return = log(amount2) - log(amount1))

# List all quarters in your dataset
all_quarters <- seq(min(c(dade$q1, dade$q2)), max(c(dade$q1, dade$q2)), by = "quarter")

# Map quarters to numeric indices
quarter_index <- setNames(seq_along(all_quarters), all_quarters)
dade <- dade %>%
  mutate(
    q1_idx = quarter_index[as.character(q1)],
    q2_idx = quarter_index[as.character(q2)]
  )

n_quarters <- length(all_quarters)
X <- matrix(0, nrow = nrow(dade), ncol = n_quarters - 1)

for (i in 1:nrow(dade)) {
  X[i, dade$q1_idx[i]:(dade$q2_idx[i]-1)] <- 1
}

y <- dade$log_return
index_coef <- solve(crossprod(X), crossprod(X, y))
index_levels <- c(0, cumsum(index_coef))  # start index at 0

data.frame(
  quarter = all_quarters, 
  index = index_levels,
  rate = c(1, exp(index_coef))
  ) %>% 
  ggplot(aes(x = quarter, y = rate)) +
  geom_line()


```


### Pissing me off

```{r}
# dade_df <- dade %>%
#   mutate(
#     qs_spanned = purrr::map2(q1, q2, ~ seq(.x, .y, by = "quarter"))
#     )
# 
# dade_df <- dade_df %>% 
#   mutate(
#     log_p1 = log(amount1),
#     log_p2 = log(amount2)
#   )

library(Matrix)

all_q <- seq(min(dade$q1), max(dade$q2), by = "quarter")
q_index <- setNames(seq_along(all_q), all_q)
nq <- length(all_q)

dade <- dade %>%
  mutate(
    i1 = q_index[as.character(q1)],
    i2 = q_index[as.character(q2)]
  )
```


```{r}
df <- dade
n <- nrow(df)

rows <- c(seq_len(n), seq_len(n))
cols <- c(df$i1, df$i2)
vals <- c(rep(-1, n), rep(1, n))

X <- sparseMatrix(
  i = rows,
  j = cols,
  x = vals,
  dims = c(n, nq)
)

y <- log(df$amount2) - log(df$amount1)
beta <- solve(crossprod(X), crossprod(X, y))
app_rates <- exp(beta) %>% as.vector()

df_lm <- as.data.frame(as.matrix(X))
fit <- lm(y ~ . -1, data = df_lm)  # no intercept

app_rates_rel <- app_rates / app_rates[1]
rate_lm_rel <- exp(coef(fit)) / exp(coef(fit)[1])

rs_rates <- data.frame(
  quarter = all_q,
  rate = app_rates,
  rate_lm = exp(coef(fit))
  # rate2 = rate2
)

rs_rates %>% 
  pivot_longer(!"quarter") %>% 
# rs_rates %>% 
  ggplot(aes(x = quarter, y = value, color = name)) +
  geom_line()


```


```{r}
library(Rlinsolve)

df <- dade %>%
  filter(strata == "S3")

n <- nrow(df)

rows <- c(seq_len(n), seq_len(n))
cols <- c(df$i1, df$i2)
vals <- c(rep(-1, n), rep(1, n))

X <- sparseMatrix(
  i = rows,
  j = cols,
  x = vals,
  dims = c(n, nq)
)
y <- log(df$amount2) - log(df$amount1)

df_lm <- as.data.frame(as.matrix(X))
fit <- lm(y ~ . - 1, data = df_lm)  # no intercept

# beta <- lsolve.bicg(crossprod(X), crossprod(X, y))
beta <- solve(crossprod(X), crossprod(X, y))

#if fails
cholX <- Cholesky(crossprod(X))
beta <- solve(cholX, crossprod(X, y))

app_rates3 <- exp(beta) %>% as.vector()

#if fails
library(RSpectra)
beta <- lsqr(X, y)$x

rs_regression <- function(df) {
  n <- nrow(df)
  rows <- c(seq_len(n), seq_len(n))
  cols <- c(df$i1, df$i2)
  vals <- c(rep(-1, n), rep(1, n))
  
  X <- sparseMatrix(
    i = rows,
    j = cols,
    x = vals,
    dims = c(n, nq)
  )
  
  y <- log(df$amount2) - log(df$amount1)
  
  beta <- solve(crossprod(X), crossprod(X, y))
  app_rates <- exp(beta)
  rs_rates <- data.frame(
    quarter = all_q,
    rate = as.vector(app_rates)
  )
  return(rs_rates)
}

strata <- paste0("S", 1:5)

strata_rates <- lapply(strata, \(s) rs_regression(filter(dade, strata == s)))

rs_rates %>% 
  pivot_longer(!"quarter") %>% 
  ggplot(aes(x = quarter, y = value, color = name)) +
  geom_line()
```

